---
title: "Multi traj RAPToR"
author: "Baptiste Rousseau"
date: "2023-11-01"
output:
  html_document:
    toc: true
    toc_depth: 2
    toc_float: true
    code_folding: hide
    code_download: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=TRUE, warning=FALSE)
```

```{r library, message=FALSE}
library(Biobase)
library(dplyr)
library(GEOquery)
library(ica)
library(limma)
library(RAPToR)
library(Seurat)
library(slingshot)
library(scales)
library(tradeSeq)
library(viridis)
library(wormRef)

source("multige_im.R")
```

```{r}
dspacker <- readRDS("./data/eset_packer.rds")
dspacker$lineage2 <- substr(dspacker$lineage, 1, 2) # make 2 char lineage

# select neurons lineages ASE, ASJ and AUA
csel <- grepl(pattern = "ASE|ASJ|AUA", dspacker$cell.subtype)
table(dspacker$cell.subtype[csel], dspacker$embryo.time.bin[csel])

pX <- exprs(dspacker) # extract counts
table(rowSums(pX) < 10) # filter low-expressed genes (also to lighten the data)

# select annotated cells from one batch
sel_cells <- dspacker$batch == "Waterston_400_minutes" & dspacker$lineage != "unannotated" & grepl(pattern = "ASE|ASJ|AUA", dspacker$cell.subtype)

pX <- as.matrix(pX[rowSums(pX) > 10, sel_cells])

p_fpx <- pData(dspacker)[sel_cells,]
```

# Preprocessing data

```{r}
# normalize & log
pX <- limma::normalizeBetweenArrays(pX, method = "quantile")
pX <- log1p(pX)

# compute correlation matrix
cc <- cor(pX, method="spearman")
diag(cc) <- NA # remove 1 diagonal
filt <- apply(cc, 1, quantile, probs=.99, na.rm=T)
thr <- mean(filt) - 2*sd(filt) # define threshold based on 99th percentile distribution
boxplot(cc, border = 1+(filt<=thr), col = 0+7*(filt<=thr), names=NA,
        xlab = "Samples", ylab = "Spearman corr. with other samples", )

pX <- pX[,filt>thr]
p_fpx <- p_fpx[filt>thr,]
```

We remove `r sum(filt<thr)` samples and keep `r sum(filt>thr)`.

```{r}
seu <- CreateSeuratObject(counts = pX)
seu <- NormalizeData(seu)
seu <- FindVariableFeatures(seu, nfeatures = 2000)
seu <- ScaleData(seu)

seu <- AddMetaData(seu, p_fpx["cell.subtype"], 
                   col.name = "cell.subtype")
```

Distribution of different celltypes

```{r, fig.height=8}
VlnPlot(seu, c("nCount_RNA", "nFeature_RNA"), pt.size = 0.1, ncol = 1, group.by = "cell.subtype")
```

# Dimension reduction

UMAP can better preserve pairwise distance of cells than tSNE and can better separate cell populations than the first 2 PCs of PCA, so we will work with UMAP rather than tSNE or PCA.

```{r, message=FALSE}
seu <- RunPCA(seu)
seu <- RunUMAP(seu, n.neighbors = 10, dims = 1:50, spread = 2, min.dist = 0.3, seed.use = 654)
DimPlot(seu, reduction = "umap",
        group.by = "cell.subtype", pt.size = 0.5, label = TRUE, repel = TRUE)
```

# Search clusters

clusters correspond to cell types

```{r, message=FALSE}
seu <- FindNeighbors(seu, verbose = FALSE, dims = 1:50)
seu <- FindClusters(seu, algorithm = 1, random.seed = 654, resolution = 1)

# Plot the clusters
DimPlot(seu, reduction = "umap", group.by = "seurat_clusters", pt.size = 0.5, label = TRUE)
```

# Slingshot

```{r}
# slingshot ----
set.seed(654)
sds <- slingshot(Embeddings(seu, "umap"), clusterLabels = seu$seurat_clusters, stretch = 0)
sds <- as.SlingshotDataSet(sds)

# create the color palette
cell_pal <- function(cell_vars, pal_fun,...) {
  if (is.numeric(cell_vars)) {
    pal <- pal_fun(100, ...)
    return(pal[cut(cell_vars, breaks = 100)])
  } else {
    categories <- sort(unique(cell_vars))
    pal <- setNames(pal_fun(length(categories), ...), categories)
    return(pal[cell_vars])
  }
}

cell_colors <- cell_pal(seu$cell.subtype, brewer_pal("qual", "Set2"))
cell_colors_clust <- cell_pal(seu$seurat_clusters, hue_pal())


curves <- as.SlingshotDataSet(getCurves(sds, approx_points = 300, thresh = 0.01,
                                        stretch = 0.8, allow.breaks = FALSE,
                                        shrink = 0.99))
```

Lineages inferred by slingshot :

```{r}
plot(slingReducedDim(sds), col = cell_colors, pch = 16)
lines(curves, lwd = 3, col = "black")
```

```{r}
pt <- slingPseudotime(sds)
nms <- colnames(pt)
pal <- viridis(100, end = 0.95)
par(mfrow = c(1, 2))
for (i in nms) {
  colors <- pal[cut(pt[,i], breaks = 100)]
  plot(slingReducedDim(sds), col = colors, pch = 16, main = i)
  lines(curves, lwd = 2, col = 'black')
}
```

# Differential expression

```{r}
counts <- as.matrix(seu@assays$RNA@counts[seu@assays$RNA@var.features, ])
dim(counts)

sce <- fitGAM(counts = as.matrix(counts), sds = curves)
```

We can look at which genes are differentially expressed along lineages.

```{r}
pseudotime_association <- associationTest(sce)
pseudotime_association$fdr <- p.adjust(pseudotime_association$pvalue, method = "fdr")
pseudotime_association <- pseudotime_association[order(pseudotime_association$pvalue), ]
pseudotime_association$feature_id <- rownames(pseudotime_association)

plot_differential_expression <- function(feature_id, counts, clustering) {
  feature_id <- pseudotime_association %>% filter(pvalue < 0.05) %>% top_n(1, -waldStat) %>% pull(feature_id)
  cowplot::plot_grid(plotGeneCount(curves, counts, gene = feature_id[1], clusters = clustering, models = sce) + ggplot2::theme(legend.position = "none"), 
                     plotSmoothers(sce, as.matrix(counts), gene = feature_id[1]))
}

feature_id <- pseudotime_association %>% filter(pvalue < 0.05) %>% top_n(1, -waldStat) %>% pull(feature_id)

plot_differential_expression(feature_id, counts, seu$seurat_clusters)
```

# Dimension reduction

```{r}
p_fpx$embryo.time <- p_fpx$embryo.time - min(p_fpx$embryo.time) + 1

cell_subtypes <- seu$cell.subtype

# for plotting
cols <- viridisLite::inferno(max(p_fpx$embryo.time), end=.9)[p_fpx$embryo.time]
pchs <- as.numeric(as.factor(cell_subtypes))
transp <- function(col, a=.5){
  # make color transparent 
  colr <- col2rgb(col)
  return(rgb(colr[1,], colr[2,], colr[3,], a*255, maxColorValue = 255))
}

ctypes <- unique(cell_subtypes)

# define lineage cell sets
lins <- list(
  ASJ = cell_subtypes %in% ctypes[c(1, 4, 5)],
  AUA = cell_subtypes %in% ctypes[c(1, 5, 6)],
  ASE = cell_subtypes %in% ctypes[c(1, 2, 3)]
)

tpX <- scale(t(pX), scale = FALSE, center = TRUE)

pca_pX <- summary(prcomp(tpX, scale. = F, center = F))
```

```{r}
mt <- multige_im(X = pX, p = p_fpx, lineages = lins,
                 dim_red = 'ica', nc = 60,
                 formulas = c("X~s(embryo.time, bs='cr', k=6)",
                              "X~s(embryo.time, bs='cr', k=6)",
                              "X~s(embryo.time, bs='cs', k=6)",
                              "X~s(embryo.time, bs='cr', k=5)",
                              "X~s(embryo.time, bs='cr', k=5)",
                              "X~s(embryo.time, bs='cr', k=5)",
                              "X~s(embryo.time, bs='cr', k=4)",
                              "X~s(embryo.time, bs='cr', k=4)"))

# predict new data (in comp. space)
newdat <- data.frame(embryo.time=seq(1,11, l=100))
nX <- predict_mgeim(mt, newdat, as.c = T)
```


```{r, message=FALSE, results=FALSE, fig.dim=c(8, 5)}
ncs <- 1:8
sapply(seq_along(mt$lineages), function(li){
  sl <- mt$lineages[[li]]
  svg(paste0("fig/neurons_test_", names(mt$lineages)[li],".svg"))
  par(mfrow = c(5,8/2), mar=c(3,1,2,1), bty='l', pty='s')
  sapply(1:ncs, function(ci){
    plot(p_fpx$embryo.time[sl], mt$dim_red$S[sl,ci], xlab = "time", ylab="IC",
         col = transp(cols, a=.2), pch = pchs,
         main = paste0(names(mt$lineages)[li], "lin, IC", mt$dim_red$ncs[ci]),
         ylim = range(mt$dim_red$S[,ci]), xlim = range(p_fpx$embryo.time))
    points(p_fpx$embryo.time[-sl], mt$dim_red$S[-sl,ci], cex = .1, col = cols[sl], pch = pchs[sl], lw=2)
    # points(newdat$blastomere_time, nX[[li]][,ci], type  ='l', col = 'red')
  })
  dev.off()
})
```


```{r, message=FALSE, results=FALSE, fig.dim=c(8, 5)}
ncs <- 1:8
par(mfrow = c(5, max(ncs)), mar=c(3,1,2,1), pty='s', bty='l')
lapply(seq_along(lins), function(li){ # for each lineage
  l <- lins[[li]]
  sapply(ncs, function(i){ # for each component
    plot(p_fpx$embryo.time, pca_pX$x[,i], 
         col = transp(cols, a=.2), pch = pchs,
         xlab = "blast stage", ylab = "PC",
         ylim=range(pca_pX$x[,i]),
         main = paste0(names(lins)[li]," lin, PC",i))
    points(p_fpx$embryo.time[l], pca_pX$x[l,i], lwd=2,
           col = cols[l], pch = pchs[l])
    
  })
})
```

# interpolation

```{r}
# GAM model on PCs
model_gam_pca <- function(X, p, formula, nc = ncol(X), subset=T){
  formula <- stats::as.formula(formula)
  
  tXc <- scale(t(X), center=T, scale = F)
  pca <- stats::prcomp(tXc, center=F, scale=F, rank=nc)
  pca$gcenters <- attr(tXc, "scaled:center")
  
  # subset PCs and p according to lineage
  pca$x <- pca$x[subset, ]
  p <- p[subset, ]
  
  formulas <- lapply(seq_len(nc), function(i) stats::update(formula, paste0("PC", i, " ~ .")))
  colnames(pca$x) <- paste0("PC", seq_len(nc))
  p <- cbind(p, pca$x)
  
  m <- list()
  m$model <- lapply(formulas, mgcv::gam, data = p)
  m$pca <- pca
  
  return(m)
}

# predict gam output
predict_gam_pca <- function(m, newdata, as.pc = FALSE){
  preds <- do.call(cbind, lapply(m$model, predict, newdata = newdata))
  
  if(!as.pc)
    return(apply(tcrossprod(m$pca$rotation, preds), 2, function(co) co + m$pca$gcenters))
  else return(preds)
}
```


```{r}
ncs <- 1:8
# build models
ms <- lapply(seq_along(lins), function(li){
  m <- model_gam_pca(X = pX, p = p_fpx, # gene expr and pdata
                     nc = max(ncs), # components to use
                     formula = "X~s(embryo.time, bs='cr', k=6)", # gam formula
                     subset = lins[[li]] # lineage subset
  )
})

# build prediction data
ndat <- data.frame(embryo.time=seq(min(p_fpx$embryo.time), max(p_fpx$embryo.time), l=100))

# predict in component space
ps_c <- lapply(ms, predict_gam_pca, newdata=ndat, as.pc=T)

# build refs (=predict in gene space)
rfs <- lapply(seq_along(ms), function(i)
  list(interpGE=predict_gam_pca(ms[[i]], ndat), 
       time=ndat$embryo.time))
```

```{r, message=FALSE, results=FALSE, fig.dim=c(8, 5)}
par(mfrow = c(5,max(ncs)), mar=c(3,1,2,1), pty='s', bty='l')
lapply(seq_along(lins), function(li){ # for each lineage
  l <- lins[[li]]
  sapply(ncs, function(i){ # for each component
    plot(p_fpx$embryo.time, pca_pX$x[,i], 
         col = transp(cols, a=.2), pch = pchs,
         xlab = "blast stage", ylab = "PC",
         ylim=range(pca_pX$x[,i]),
         main = paste0(names(lins)[li]," lin, PC",i))
    # show lineage-specific points 
    points(p_fpx$embryo.time[l], pca_pX$x[l,i], lwd=2,
           col = cols[l], pch = pchs[l])
    
    # add interpolation
    points(ndat$embryo.time, ps_c[[li]][,i], lwd=2, type = "l", col='green')
  })
})
```

# Staging

```{r, message=FALSE, results=FALSE}
ae_s <- lapply(rfs, function(r){
  RAPToR::ae(pX, refdata = r$interpGE, ref.time_series = r$time, bootstrap.n = 30)
})

par(mfrow = c(3,3), mar=c(3,1,2,1), pty='s', bty='l')
xl <- range(p_fpx$embryo.time)
sapply(seq_along(lins), function(i){
  l <- lins[[i]]
  sapply(seq_along(lins), function(j){
    x <- p_fpx$embryo.time[l]
    y <- ae_s[[j]]$age.estimates[l,1]
    
    plot(x, y, xlim = xl, ylim = xl,
         main = paste0(names(lins)[i], ' lin /', names(lins)[j], ' ref'), 
         col = i, xlab = "Blast. time", ylab="Estimated age")
    
    abline(a=0, b=1, lty=2) # add x=y
    # add cor.scores
    mtext(text = paste0("r = ", round(cor(x, y, method = 'pearson'), 3),
                        "\nrho =", round(cor(x, y, method = 'spearman'), 3)),
          at = 1, side = 3, line = -2, cex = .8, adj = 0)
    
  })
})
```

# Exploring lineage-inference from staging

```{r, message=FALSE, results=FALSE}
par(mfrow = c(2,3), mar=c(3,3,2,1), bty='l')
sapply(ctypes[c(1,5,4,6,3,2)], function(ci){
  ss <- which(cell_subtypes==ci) # select a cell type
  
  # get cor. score at estimate for each reference
  ccs <- do.call(cbind, lapply(ae_s, function(a) a$age.estimates[ss,"cor.score"]))
  colnames(ccs) <- names(lins)
  boxplot(ccs, border=seq_along(lins), las=2,
          lwd=2, ylim=c(0, 1), col = 0, boxwex=.4,
          ylab = "Cor. at estimate", xlab = "reference used", main = ci)
  mtext(paste0('n=',length(ss)), at=.5, line=-1, cex=.75, side=3, adj = 0)
})
```

# Integrate more data

We will now integrate the 40th first dimensions of the PCA, counting for 90% of the explained variance.

```{r, message=FALSE, results=FALSE}
ncs <- 1:40 # 90% of varexp
# build models
ms2 <- lapply(seq_along(lins), function(li){
  m <- model_gam_pca(X = pX, p = p_fpx, # gene expr and pdata
                     nc = max(ncs), # components to use
                     formula = "X~s(embryo.time, bs='cr', k=6)", # gam formula
                     subset = lins[[li]] # lineage subset
  )
})

# build refs (=predict in gene space)
rfs2 <- lapply(seq_along(ms), function(i)
  list(interpGE=predict_gam_pca(ms2[[i]], ndat), 
       time=ndat$embryo.time))

ae_s2 <- lapply(rfs2, function(r){
  RAPToR::ae(pX, refdata = r$interpGE, ref.time_series = r$time, bootstrap.n = 30)
})
```

```{r, message=FALSE, results=FALSE}
par(mfrow = c(2,3), mar=c(3,3,2,1), bty='l')
sapply(ctypes[c(1,5,4,6,3,2)], function(ci){
  ss <- which(cell_subtypes==ci) # select a cell type
  
  # get cor. score at estimate for each reference
  ccs <- do.call(cbind, lapply(ae_s2, function(a) a$age.estimates[ss,"cor.score"]))
  colnames(ccs) <- names(lins)
  boxplot(ccs, border=seq_along(lins), las=2,
          lwd=2, ylim=c(0, 1), col = 0, boxwex=.4,
          ylab = "Cor. at estimate", xlab = "reference used", main = ci)
  mtext(paste0('n=',length(ss)), at=.5, line=-1, cex=.75, side=3, adj = 0)
  
})
```

# Normalize

Since we only want to observe differences in correlation between lineages, we normalize for each cell by subtracting the smallest correlation score between lineages. This way, we remove the effect of sample quality or stage/age on the base correlation score, and only observe the effect of using different lineage references.

First with the 8th first dimensions.

```{r, message=FALSE, results=FALSE}
# function to normalize by min value in series
levelmin <- function(x){(x-min(x))}

par(mfrow = c(2,3), mar=c(3,2,2,1), bty='l', pty='s')
sapply(ctypes[c(1,5,4,6,3,2)], function(ci){
  ss <- which(cell_subtypes==ci) # select a cell type
  
  # get cor. score at estimate for each reference
  ccs <- do.call(cbind, lapply(ae_s, function(a) a$age.estimates[ss,"cor.score"]))
  ccs <- t(apply(ccs,1, levelmin))
  x <- p_fpx$embryo.time[ss]
  colnames(ccs) <- names(lins)
  plot(range(p_fpx$embryo.time), range(ccs), type = 'n',
       lwd=2, ylim=c(0, .05), col = 0,
       ylab = "Correlation diff. between lineages", xlab = "Blast. time", main = ci)
  sapply(seq_along(lins), function(i){
    points(x, ccs[,i], col = transp(i, a = .9), lwd=2)
  })
  if(ci==ctypes[1])
    legend('right', bty='n', legend=names(lins), 
           lwd=3, lty=NA, pch=1, col=seq_along(lins),
           title = "Reference")
})
```

And then with the 40.

```{r, message=FALSE, results=FALSE}
par(mfrow = c(2,3), mar=c(3,2,2,1), bty='l', pty='s')
sapply(ctypes[c(1,5,4,6,3,2)], function(ci){
  ss <- which(cell_subtypes==ci) # select a cell type
  
  # get cor. score at estimate for each reference
  ccs <- do.call(cbind, lapply(ae_s2, function(a) a$age.estimates[ss,"cor.score"]))
  ccs <- t(apply(ccs,1, levelmin))
  x <- p_fpx$embryo.time[ss]
  colnames(ccs) <- names(lins)
  plot(range(p_fpx$embryo.time), range(ccs), type = 'n',
       lwd=2, ylim=c(0, .11), col = 0,
       ylab = "Correlation diff. between lineages", xlab = "Blast. time", main = ci)
  sapply(seq_along(lins), function(i){
    points(x, ccs[,i], col = transp(i, a = .9), lwd=2)
  })
  if(ci==ctypes[1])
    legend('right', bty='n', legend=names(lins), 
           lwd=3, lty=NA, pch=1, col=seq_along(lins),
           title = "Reference")
})
```
