---
title: "Multi traj RAPToR"
author: "Baptiste Rousseau"
date: "2023-11-01"
output:
  html_document:
    toc: true
    toc_depth: 2
    toc_float: true
    code_folding: hide
    code_download: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=TRUE, warning=FALSE)
```

```{r library, message=FALSE}
library(Biobase)
library(dplyr)
library(GEOquery)
library(ica)
library(limma)
library(RAPToR)
library(Seurat)
library(slingshot)
library(scales)
library(tradeSeq)
library(viridis)
library(wormRef)

source("ale.R")
source("multige_im.R")
source("utils.R")
```

Load data and keep only neurons lineages.

```{r}
dspacker <- readRDS("./data/eset_packer.rds")
dspacker$lineage2 <- substr(dspacker$lineage, 1, 2) # make 2 char lineage

# select neurons lineages ASE, ASJ and AUA
csel <- grepl(pattern = "ASE|ASJ|AUA", dspacker$cell.subtype)
table(dspacker$cell.subtype[csel], dspacker$embryo.time.bin[csel])

pX <- exprs(dspacker) # extract counts
```

# Preprocessing data

We remove non/low-expressed genes and quantile-normalize the expression data, as well as apply a log(X+1) transformation. We then filter out poor quality samples using correlation with other samples.

```{r}
table(rowSums(pX) < 10) # filter low-expressed genes (also to lighten the data)

# select annotated cells from one batch
sel_cells <- dspacker$batch == "Waterston_400_minutes" & dspacker$lineage != "unannotated" & grepl(pattern = "ASE|ASJ|AUA", dspacker$cell.subtype)

pX <- as.matrix(pX[rowSums(pX) > 10, sel_cells])

p_fpx <- pData(dspacker)[sel_cells,]

# normalize & log
pX <- limma::normalizeBetweenArrays(pX, method = "quantile")
pX <- log1p(pX)

# compute correlation matrix
cc <- cor(pX, method="spearman")
diag(cc) <- NA # remove 1 diagonal
filt <- apply(cc, 1, quantile, probs=.99, na.rm=T)
thr <- mean(filt) - 2*sd(filt) # define threshold based on 99th percentile distribution
boxplot(cc, border = 1+(filt<=thr), col = 0+7*(filt<=thr), names=NA,
        xlab = "Samples", ylab = "Spearman corr. with other samples", )

pX <- pX[,filt>thr]
p_fpx <- p_fpx[filt>thr,]
```

We remove `r sum(filt<thr)` samples and keep `r sum(filt>thr)`.

```{r}
seu <- CreateSeuratObject(counts = pX)
seu <- NormalizeData(seu)
seu <- FindVariableFeatures(seu, nfeatures = 2000)
seu <- ScaleData(seu)

seu <- AddMetaData(seu, p_fpx["cell.subtype"], 
                   col.name = "cell.subtype")
```

Distribution of different cell types :

```{r, fig.height=8}
VlnPlot(seu, c("nCount_RNA", "nFeature_RNA"), pt.size = 0.1, ncol = 1, group.by = "cell.subtype")
```

# Dimension reduction

UMAP can better preserve pairwise distance of cells than tSNE and can better separate cell populations than the first 2 PCs of PCA, so we will work with UMAP for visualization rather than tSNE or PCA.

```{r, message=FALSE}
seu <- RunPCA(seu)
seu <- RunUMAP(seu, n.neighbors = 10, dims = 1:50, spread = 2, min.dist = 0.3, seed.use = 654)
DimPlot(seu, reduction = "umap",
        group.by = "cell.subtype", pt.size = 0.5, label = TRUE, repel = TRUE)
```

```{r}
p_fpx$embryo.time <- p_fpx$embryo.time - min(p_fpx$embryo.time) + 1

cell_subtypes <- seu$cell.subtype

ctypes <- unique(cell_subtypes)

# define lineage cell sets
lins <- list(
  ASJ = cell_subtypes %in% ctypes[c(1, 4, 5)],
  AUA = cell_subtypes %in% ctypes[c(1, 5, 6)],
  ASE = cell_subtypes %in% ctypes[c(1, 2, 3)]
)
```

```{r message=FALSE, results=FALSE}
prepare_and_interpolate <- function(X, p, lins, dim_red="ica", nc=8){
  mt <- multige_im(X = X, p = p, lineages = lins,
                   dim_red = dim_red, nc = nc,
                   formulas = c("X~s(embryo.time, bs='cr', k=6)"))
  
  # predict new data (in comp. space)
  ndat <- data.frame(embryo.time=seq(min(p$embryo.time),
                                     max(p$embryo.time), l=100))
  nX <- predict_mgeim(mt, ndat, as.c = T)
  interpGE <- predict_mgeim(mt, ndat)
  
  return(list(mt=mt, ndat=ndat, nX=nX, interpGE=interpGE))
}
```

# RAPToR

```{r}
# build train and test
p <- 0.8
rr <- split(1:length(cell_subtypes), cell_subtypes)
idx <- sort(as.numeric(unlist(sapply(rr, function(x) sample(x, length(x) * p)))))

pX_train <- pX[idx, ]
pX_test <- pX[-idx, ]

p_fpx_train <- p_fpx[idx, ]
p_fpx_test <- p_fpx[-idx, ]
```

First let's build sets with 80/20 while keeping original proportions. In other words, stratified splitting of the dataset.

Original proportions :
```{r}
table(p_fpx$cell.subtype) / nrow(p_fpx)
```

Train proportions :
```{r}
table(p_fpx_train$cell.subtype) / nrow(p_fpx_train)
```

Test proportions :
```{r}
table(p_fpx_test$cell.subtype) / nrow(p_fpx_test)
```

We can see that the proportions are similar enough.

# Interpolation

```{r, message=FALSE, results=FALSE}
multige <- prepare_and_interpolate(pX, p_fpx, lins)
```

```{r, message=FALSE, results=FALSE}
pairs(multige$mt$dim_red$S[,1:4], col = cols, pch=pchs, lwd=2, labels=paste0("Comp.", 1:4), oma=c(2,2,2,25))
par(xpd = TRUE)
legend("bottomright", pch = as.numeric(as.factor(unique(cell_subtypes))),
       legend = c(levels(as.factor(cell_subtypes))), title="Lineage")
legend_image <- as.raster(matrix(viridisLite::inferno(max(p_fpx$embryo.time), end=.9), ncol=1))
text(x=.85, y=.85, labels = "Time")
rasterImage(legend_image, .8, .8, .9, .5)
rect(.8, .5, .9, .8)
labs <- seq(0, max(p_fpx$embryo.time),50)
ys <- seq(.5, .78, l=length(labs))
text(x=.903, y=ys, labels = "-") 
text(x=.94, y=ys, labels = labs)
```

```{r, message=FALSE, results=FALSE}
plot_lineages(multige$mt, multige$ndat, multige$nX,
              p_fpx$embryo.time, p_fpx$cell.subtype, 1:8)
```

```{r, message=FALSE, results=FALSE}
res <- lapply(1:4, function(i){
  set.seed(i)
  p <- 0.8
  idx <- stratified_split(cell_subtypes, p)
  
  pX_train <- pX[, idx]
  pX_test <- pX[, -idx]
  
  p_fpx_train <- p_fpx[idx, ]
  p_fpx_test <- p_fpx[-idx, ]
  
  lins_train <- lapply(lins,'[', idx)
  lins_test <- lapply(lins,'[', -idx)
  
  multige <- prepare_and_interpolate(pX_train, p_fpx_train, lins_train)
  
  ae_s_train <- ale(samp = pX_train, multiref=list(interpGE=multige$interpGE, time=multige$ndat$embryo.time))
  
  ae_s_test <- ale(samp = pX_test, multiref=list(interpGE=multige$interpGE, time=multige$ndat$embryo.time))
  
  return(list(multige=multige,
              train=list(idx=idx,
                         ae_s=ae_s_train),
              test=list(idx=seq_along(cell_subtypes)[!seq_along(cell_subtypes) %in% idx],
                        ae_s=ae_s_test)))
})
```

# Staging

Predictions for train :

```{r, message=FALSE, results=FALSE}
plot_staged(res[[1]]$train$ae_s$ae.lin, p_fpx[res[[1]]$train$idx, ]$embryo.time,
            lapply(lins,'[', res[[1]]$train$idx))
```

Predictions for test :

```{r, message=FALSE, results=FALSE}
lapply(seq_along(res), function(i){
  plot_staged(res[[i]]$test$ae_s$ae.lin, p_fpx[res[[i]]$test$idx, ]$embryo.time,
              lapply(lins,'[', res[[i]]$test$idx), mode="diag")
})
```

Results from 4 differents runs :

train pearson r :
```{r}
reduce_cor(res, lins, p_fpx, "train", "pearson")
```

train spearman rho :
```{r}
reduce_cor(res, lins, p_fpx, "train", "spearman")
```

test pearson r :
```{r}
reduce_cor(res, lins, p_fpx, "test", "pearson")
```

test spearman rho :
```{r}
reduce_cor(res, lins, p_fpx, "test", "spearman")
```

# Exploring lineage-inference from staging

```{r, message=FALSE, results=FALSE}
plot_lineages_inference(res[[1]]$train$ae_s$ae.lin,
                        p_fpx[res[[1]]$train$idx, ]$cell.subtype,
                        ctypes[c(1,5,4,6,3,2)],
                        lapply(lins,'[', res[[1]]$train$idx))
```


```{r, message=FALSE, results=FALSE}
plot_lineages_inference(res[[1]]$test$ae_s$ae.lin,
                        p_fpx[res[[1]]$test$idx, ]$cell.subtype,
                        ctypes[c(1,5,4,6,3,2)],
                        lapply(lins,'[', res[[1]]$test$idx))
```

Results from 4 different runs :

train :
```{r}
reduce_summary(res, lins, ctypes[c(1,5,4,6,3,2)], "train")
```

test :
```{r}
reduce_summary(res, lins, ctypes[c(1,5,4,6,3,2)], "test")
```

# Normalization

Predictions for train :

```{r, message=FALSE, results=FALSE}
plot_lineage_inference_norm(res[[1]]$train$ae_s$ae.lin,
                            p_fpx[res[[1]]$train$idx, ]$cell.subtype,
                            ctypes[c(1,5,4,6,3,2)],
                            p_fpx[res[[1]]$train$idx, ]$embryo.time,
                            lapply(lins,'[', res[[1]]$train$idx))
```

Predictions for test :

```{r, message=FALSE, results=FALSE}
plot_lineage_inference_norm(res[[1]]$test$ae_s$ae.lin,
                            p_fpx[res[[1]]$test$idx, ]$cell.subtype,
                            ctypes[c(1,5,4,6,3,2)],
                            p_fpx[res[[1]]$test$idx, ]$embryo.time,
                            lapply(lins,'[', res[[1]]$test$idx))
```

Results from 4 different runs :

train :
```{r}
reduce_lineage_inference(res, p_fpx, "train")
```

test :
```{r}
reduce_lineage_inference(res, p_fpx, "test")
```

